A preliminary constraint grammar for Russian

### 2015-05-10 2015-nodalida-ruscg/presentation_ruscg.tex
revisions
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60326 72bbbca6-d526-0410-a7d9-f06f51895060

\begin{frame}
\frametitle{Ambiguity in Russian}
\framesubtitle{}
\begin{itemize}
	\item Russian is characterized by fusional morphology
	\begin{itemize}
		\item Nouns: 12+ paradigm cells (6 main cases, sg and pl)
		\item Modifiers: 34+ paradigm cells (6 cases, 3 genders, etc.)
		\item Verbs: 13+ paradigm cells (4 past, 6 non-past, imperatives, etc.)
		% 200+ if you count participles

		\item \rus{выпей} \emph{vypej} `drink.IMP' or `bittern.Pl-GEN'
	\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Frequency of each kind of ambiguity}
\framesubtitle{}
\begin{table}
  \centering
  \begin{tabular}{l|rr}
    \hline
    \textbf{Type}  & \textbf{all tokens} & \textbf{ambig. tokens} \\
    \hline
    Intraparadigm. & 59.0\%                   & 90.9\%   \\
    Incongruent    & 27.7\%                   & 42.7\%   \\ 
    Congruent      & 1.2\%                   & 1.8\%    \\ 
    \hline
  \end{tabular}
  \caption{Frequency of different types of morphosyntactic ambiguity in unrestricted text}
  %TODO caption says 'Table :' without a number
  \label{table:ambiguity}
\end{table}
\begin{itemize}
	\item \rus{з\'{a}мков} \emph{zámkov} `castle.PL-GEN'
	\item \rus{замк\'{о}в} \emph{zamkóv} `lock.PL-GEN'
	\item \rus{з\'{а}мков} \emph{zámkov} `having to do with castles.A-MSC-SG-PRED'
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{CG rules: distribution of rule across levels}
\framesubtitle{}
\begin{table}
  \centering
  \begin{tabular}{lrrr}
    \hline
                     & {\sc select} & {\sc remove} & {\sc map} \\
    \hline
    Safe             &   16         &   34         &  -- \\
    Safe heuristic   &   89         &   76         &  -- \\
    Heuristic        &   26         &   52         &  -- \\
    Syntax labelling & --           & --           & 6 \\ 
    \hline
  \end{tabular}
  \caption{The 299 rules in the grammar are separated into four sections depending
      on rule reliability. }
  \label{tab:ruleDist}
\end{table}
\begin{itemize}
	\item Safe: real constraints in the language\\
		{\small Ex: preposition cannot directly precede a finite verb}
		\pause
	\item Safe heuristic: highly reliable tendencies\\ % Russian word order is very free
		{\small Ex: remove GEN from sentence-init. if no verb governing GEN.}
		\pause
	\item Heuristic: tendencies that are true most of the time\\ % better replaced with machine-learning
		{\small Ex: \rus{такая} is `such.FEM-SG-NOM', not `say~\rus{так}.V-PRS-ACT-ADV'}
\end{itemize}
\end{frame}

\section{Development process} %FRAN

\begin{frame}
\frametitle{Writing rules and level assignment}
\framesubtitle{}
\begin{itemize}
	\item No hand-annotated corpus was compatible with our analyser
	\pause
	\item We sorted new rules using Russian Wikipedia by collecting 100 example applications of the rule
	\pause
	\begin{itemize}
		\item Safe: 100\% correct
		\item Safe heuristic: > 97\% correct
		\item Heuristic (or discard): $\leq$ 97\% correct 
	\end{itemize}
\end{itemize}
\end{frame}

\section{Evaluation}
\subsection{Qualitative evaluation} %FRAN
\begin{frame}
\frametitle{}
\framesubtitle{}
\begin{itemize}
	\item 
	\pause
	\item 
	\begin{itemize}
		\item 
		\pause
	\end{itemize}
	\item 
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Automatic word stress placement - results}
\framesubtitle{}
\begin{table}
  \centering
  \begin{tabular}{r | r r r }
     & accuracy\% & error\% & abstention\%  \\
    \hline
    \hline
    {\small {\tt RussianGram}} & 90.09 & 0.79 & 9.12 \\
    {\small {\tt no CG}} & 90.07 & 0.49 & 9.44  \\
    {\small {\tt CG}} & 93.21 & 0.74 & 6.05
  \end{tabular}
  \caption{Results of stress placement task evaluation}
  \label{tab:results}
\end{table}
\begin{itemize}
	\item CG makes gains of 3.14\% (42\% of ambiguity from FST)
	\pause
	\item CG's high recall leads to a low error rate
	\pause
	\item To be continued.... (tomorrow)
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Combining CG with HMM tagger}
\framesubtitle{}
\begin{itemize}
	\item CG removes less than half of ambiguity
	\pause
	\item To resolve remaining ambiguity, we backoff to HMM tagger
	\begin{itemize}
		\item This approach has been applied with...\\
		Basque \cite{ezeiza.ea-98},\\
		Czech \cite{hajic.ea-01,hajic.ea-07},\\
		Norwegian \cite{johannessen.ea-11,johannessen.ea-12},\\
		Spanish \cite{hulden12},\\
		and Turkish \cite{oflazer.tur-96}
		\pause
	\end{itemize}
	\item We follow the voting method used by \cite{hulden12}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Voting setup}
\framesubtitle{}
\begin{itemize}
	\item Freely available hunpos tagger
	\pause
	\item 10-fold cross-validation (90\% training, 10\% testing)
	\pause
	\item 3 configurations
	\begin{itemize}
		\item HMM - hunpos with default options
		\pause
		\item HMM+Morph - incorporating FST output as full-form lexicon
		\pause
		\item HMM+Morph vs CG - Voting setup: if constraint grammar left 1 valid reading,
		take that. Otherwise, take output of HMM+Morph.
	\end{itemize}
	\item 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Voting results}
\framesubtitle{}
\begin{figure}
  \centering
  \includegraphics[width=0.70\textwidth]{graphics/learning-curve.pdf}
  %\caption{Learning curve for three taggers, \texttt{hunpos} with no lexicon, 
  %  \texttt{hunpos} with a lexicon, and \texttt{hunpos} with a lexicon and the 
  %  Russian constraint grammar in a voting set up.}
  \label{fig:curve}
\end{figure}
\begin{itemize}
	\item hunpos tagger alone is more dependent on training-corpus size
	\pause
	\item 
	\pause
	\item 
	\item 
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Future work}
\framesubtitle{}
\begin{itemize}
	\item Improve precision while maintaining high recall
	\pause
	\item Add syntactic functions and dependency parsing
	\begin{itemize}
		\item We plan to use Giellatekno's dependency parser \cite{antonsen10}
		\pause
	\end{itemize}
	\item Improve workflow by extending our gold-standard corpus while checking new rule accuracy
\end{itemize}
\end{frame}

>>>>

\begin{frame}
\frametitle{Ambiguity in Russian}
\framesubtitle{}
\begin{itemize}
	\item Russian is characterized by fusional morphology
	\begin{itemize}
		\item Nouns: 12+ paradigm cells (6 main cases, sg and pl)
		\item Modifiers: 34+ paradigm cells (6 cases, 3 genders, etc.)
		\item Verbs: 13+ paradigm cells (4 past, 6 non-past, imperatives, etc.)
		% 200+ if you count participles
	\end{itemize}
	\pause
	\item Both in-category and out-of-category ambiguity
	\begin{itemize}
		\item \rus{выпей} \emph{vypej} `drink.V-IMPER' or `bittern.N-PL-GEN'
		\includegraphics[width=0.40\textwidth]{graphics/vypej3103.jpeg} %TODO add image
	\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Frequency of each kind of ambiguity}
\framesubtitle{}
\begin{table}
  \centering
  \begin{tabular}{l|rr}
    \hline
    \textbf{Type}  & \textbf{all tokens} & \textbf{ambig. tokens} \\
    \hline
    Intraparadigm. & 59.0\%                   & 90.9\%   \\
    Incongruent    & 27.7\%                   & 42.7\%   \\ 
    Congruent      & 1.2\%                   & 1.8\%    \\ 
    \hline
  \end{tabular}
  \caption{Frequency of different types of morphosyntactic ambiguity in unrestricted text}
  %TODO caption says 'Table :' without a number
  %TODO numbers different from ERC
  \label{table:ambiguity}
\end{table}
\pause
\begin{itemize}
	\item Categories are not mutually exclusive
	\begin{itemize}
		\item \rus{з\'{a}мков} \emph{zámkov} `castle.PL-GEN'
		\item \rus{замк\'{о}в} \emph{zamkóv} `lock.PL-GEN'
		\item \rus{з\'{а}мков} \emph{zámkov} `having to do with castles.A-MSC-SG-PRED'
	\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{CG rules: distribution of rule across levels}
\framesubtitle{}
\begin{table}
  \centering
  \begin{tabular}{lrrr}
    \hline
                     & {\sc select} & {\sc remove} & {\sc map} \\
    \hline
    Safe             &   16         &   34         &  -- \\
    Safe heuristic   &   89         &   76         &  -- \\
    Heuristic        &   26         &   52         &  -- \\
    Syntax labelling & --           & --           & 6 \\ 
    \hline
  \end{tabular}
  \caption{The 299 rules in the grammar are separated into four sections depending
      on rule reliability. }
  \label{tab:ruleDist}
\end{table}
\begin{itemize}
	\item Safe: real constraints in the language\\
		{\small Ex: preposition cannot directly precede a finite verb}
		\pause
	\item Safe heuristic: highly reliable tendencies\\ % Russian word order is very free
		{\small Ex: remove GEN from sentence-init. if no verb governing GEN.}
		\pause
	\item Heuristic: tendencies that are true most of the time\\ % better replaced with machine-learning
		{\small Ex: \rus{такая} is `such.FEM-SG-NOM', not `say~\rus{так}.V-PRS-ACT-ADV'}
		%TODO better example
\end{itemize}
\end{frame}

\subsection{Development process} %FRAN

\begin{frame}
\frametitle{Writing rules and level assignment}
\framesubtitle{}
\begin{itemize}
	\item No hand-annotated corpus was compatible with our analyser
	\pause
	\item We sorted new rules using Russian Wikipedia by collecting 100 example applications of the rule
	\pause
	\begin{itemize}
		\item Safe: 100\% correct
		\item Safe heuristic: > 97\% correct
		\item Heuristic (or discard): $\leq$ 97\% correct
		%TODO put before asdf
	\end{itemize}
\end{itemize}
\end{frame}

\section{Evaluation}

\begin{frame}
\frametitle{Results}
\framesubtitle{}
\begin{table}
  \centering
  \begin{tabular}{l|rrrrr}
    %\hline
    \textbf{Domain} & \textbf{Tokens} & \textbf{Precision} & \textbf{Recall} & \textbf{F-score} & {Ambig. solved} \\
    \hline
    Wikipedia       & 7,857      & 0.506        & 0.996    & 0.671 & 44.92\%  \\
    Literature      & 1,652      & 0.473        & 0.984    & 0.638 & 42.95\%  \\
    News            & 642        & 0.471        & 0.990    & 0.638 & 41.60\%  \\
    \hline
    \textbf{Total}  & 10,150     & 0.498        &  0.994   & 0.663 & 44.39\% \\
    %\hline
  \end{tabular}
  \caption{Results on the test corpora.}
  \label{table:results}
\end{table}
\end{frame}

\subsection{Qualitative evaluation} %FRAN

\begin{frame}
\frametitle{Errors from `bad linguistics' (too limiting)}
\framesubtitle{}
\begin{itemize}
	  \item In some cases a rule did not take into account grammatical possibilities 
    in the language. e.g. Two simple rules such as\ldots
    \begin{itemize}
      \item \texttt{REMOVE Det IF (0 Det OR Pron) (1C Ne) ;}
      \item \texttt{REMOVE Det IF (0 Det OR Pron) (1 Cm LINK 1 CC OR CS) ;}
    \end{itemize}
    \ldots did not take into account the possibility of having a postposed determiner as in\ldots
    \begin{itemize}
      \item \ldots \rus{а может быть и раньше, и факт этот не раз поражал меня} \ldots
      \item \ldots and maybe even earlier, and fact \textbf{this} not once surprised me \ldots
    \end{itemize}
    \ldots or a interposed parenthetical as in\ldots
    \begin{itemize}
      \item \rus{Но какие, однако же, два разные создания, точно обе с двух разных планет!}
      \item But what, \textbf{exactly} , two different creatures, just both from two different planets!
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Errors from bad rules}
\framesubtitle{}
\begin{itemize}
	\item In some cases a rule was simply incorrectly specified. For example, the 
	following rule was designed to solve the ambiguity between short-form neuter adjectives and 
	adverbs
   \begin{itemize}
		\item \texttt{REMOVE A + Short IF (-1C Fin OR Adv OR A) (0C Short OR Adv) ;}
   \end{itemize}
   \item However there is no reason why we should prefer an adverb over an adjective after an adverb, 
   \begin{itemize}
     \item \ldots \rus{потому что совсем неприятно проснуться в гробу под землею.}
     \item \ldots because [it is] really \textbf{unpleasant} to wake up in a coffin under the ground.
   \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Errors from incomplete barrier}
\framesubtitle{}
\begin{itemize}
	\item Some rules suffered from incomplete barriers, which is something that would
	benefit from a more systematic treatment.
	\begin{itemize} 
		\item \texttt{REMOVE NGDAIP - Acc - Prp - Loc IF (-1C* Pr/V OR Pr/Na BARRIER (*) - Adv - Comp - DetIndecl - ModAcc - ModPrp) ;}
	\end{itemize}
	here the rule removes the nominative reading of the adjective to leave the accusative 
	reading because the preposition \rus{в} `in' is found preceding. 
	\begin{itemize}
     \item \rus{В 1960-х электрифицированные высокоскоростные железные дороги появились в Японии и некоторых других странах.}
     \item In the 1960's \textbf{electrified} high-speed railways appeared in Japan and some other countries.
   \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Errors from incomplete sets}
\framesubtitle{}
\begin{itemize}
	\item In some cases the rule was a good generalisation, but made use of a set
	which was incomplete. For example:
	\begin{itemize}
		\item \texttt{REMOVE Dat IF (NOT 0 Prn/Sebe) (NOT 0 Anim OR Cog OR Ant) (NOT 0 Pron) (NOT 1* V/Dat) (NOT -1* V/Dat) (NOT -1* Prep/Dat) (NOT -1C A + Dat) ;}
	\end{itemize} 
	the set \texttt{V/Dat} does not contain the verb \rus{противопоставлять} `opposed to' which takes 
	a dative argument.
	\begin{itemize}
		\item \rus{В связи с этим ортодоксальности стали противопоставлять ересь.}
		\item In connection with this \textbf{orthodoxy} was opposed to heresy. 
	\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Errors from rule interaction}
\framesubtitle{}
\begin{itemize}
	\item The strong accusative rule below causes incorrect behaviour in the rule to remove transitivity readings
	\begin{itemize}
		\item \texttt{REMOVE TV - Pass IF (NOT 1* Acc) (NOT -1* Acc) ;}
		\item \texttt{REMOVE Acc IF (-1C Fin + IV) (NOT 0 AccAdv) ;}
	\end{itemize}
	Consider the following example where \rus{может} `can' is tagged as intransitive, the second rule fires removing 
	the accusative reading of \rus{его} `him', and thus given the lack of accusative reading, \rus{найти} `find'
	is disambiguated as intransitive instead of transitive.
 	\begin{itemize}
		\item \rus{Она смотрит везде, но не может его найти.}
		\item She looks around, but she cannot \textbf{find him}.
	\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Errors from difficult linguistics}
\framesubtitle{}
\begin{itemize}
	\item Dealing with participles with arguments is challenging in the case that the arguments 
	of the participle share the same government as the main verb.
	\begin{itemize}
		\item \texttt{REMOVE IV IF (0 TV OR IV) (1C Acc) (NOT 1 AccAdv);}
	\end{itemize}
		Here \rus{Ваню и Машу} `Vanja and Maša' are the object of \rus{видит} `sees' and not \rus{играющих} `playing', although
		both verbs can take accusative object.
	\begin{itemize}
		\item \rus{Их мама внутри дома с кошкой, она смотрит в окно и видит играющих Ваню и Машу.}
		\item Their mother is inside the house with the cat, she looks through the window and sees Vanja and Maša \textbf{playing}.
	\end{itemize}
	This kind of error would ideally be resolved with semantic knowledge.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Automatic word stress placement - results}
\framesubtitle{}
\begin{table}
  \centering
  \begin{tabular}{r | r r r }
     & accuracy\% & error\% & abstention\%  \\
    \hline
    \hline
    {\small {\tt RussianGram}} & 90.09 & 0.79 & 9.12 \\
    {\small {\tt no CG}} & 90.07 & 0.49 & 9.44  \\
    {\small {\tt CG}} & 93.21 & 0.74 & 6.05
  \end{tabular}
  \caption{Results of stress placement task evaluation}
  \label{tab:results}
\end{table}
\begin{itemize}
	\item CG makes gains of 3.14\% (42\% of ambiguity from FST)
	\pause
	\item CG's high recall leads to a low error rate
	\begin{itemize}
		\item Higher accuracy was achieved (96.15\%), but with higher error rate (3.14\%)
	\end{itemize}
	\pause
	\item To be continued.... (tomorrow)
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Combining CG with HMM tagger}
\framesubtitle{}
\begin{itemize}
	\item CG removes less than half of ambiguity
	\pause
	\item To resolve remaining ambiguity, we turn to stochastic methods
	\begin{itemize}
		\item This approach has been applied with...\\
		Basque \cite{ezeiza.ea-98},\\
		Czech \cite{hajic.ea-01,hajic.ea-07},\\
		Norwegian \cite{johannessen.ea-11,johannessen.ea-12},\\
		Spanish \cite{hulden12},\\
		and Turkish \cite{oflazer.tur-96}
		\pause
	\end{itemize}
	\item We follow the voting method used by \cite{hulden12}, using the freely available hunpos tagger
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Voting setup}
\framesubtitle{}
\begin{itemize}
	\item 3 configurations
	\begin{itemize}
		\item HMM - hunpos with default options
		\pause
		\item HMM+Morph - incorporating FST output as full-form lexicon
		\pause
		\item HMM+Morph vs CG - Voting setup: if constraint grammar left 1 valid reading,
		take that. Otherwise, take output of HMM+Morph.
	\end{itemize}
	\pause
	\item 10-fold cross-validation
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Voting results}
\framesubtitle{}
\begin{figure}
  \centering
  \includegraphics[width=0.70\textwidth]{graphics/learning-curve.pdf}
  %\caption{Learning curve for three taggers, \texttt{hunpos} with no lexicon, 
  %  \texttt{hunpos} with a lexicon, and \texttt{hunpos} with a lexicon and the 
  %  Russian constraint grammar in a voting set up.}
  \label{fig:curve}
\end{figure}
\begin{itemize}
	\item hunpos tagger alone is more dependent on training-corpus size
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Future work}
\framesubtitle{}
\begin{itemize}
	\item Improve precision (while maintaining high recall)
	\pause
	\item Add syntactic functions and dependency parsing
	\begin{itemize}
		\item We plan to use Giellatekno's dependency parser \cite{antonsen10}
		\pause
	\end{itemize}
	\item Improve workflow by extending our gold-standard corpus while checking new rule accuracy
	\begin{itemize}
		\item Can be used for regression testing
	\end{itemize}
\end{itemize}
\end{frame}
